---
title: "Business Intelligence Lab Submission Markdown"
author: "<trevor ngugi>"
date: "<27/11/2023>"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Student ID Numbers and Names of Group Members** | |
|                                                   |                                                                                                                                                                          |
|                                                   | 1.  133996 - B - Trevor ngugi                                                                                                                                              |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                               |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                             |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                               |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                               |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **GitHub Classroom Group Name**                   | champions                                                                                                       |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Course Code**                                   | BBT4206                                                                                                                                                                  |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Course Name**                                   | Business Intelligence II                                                                                                                                                 |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Program**                                       | Bachelor of Business Information Technology                                                                                                                              |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Semester Duration**                             | 21^st^ August 2023 to 28^th^ November 2023                                                                                                                               |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

# Setup Chunk

We start by installing all the required packages

```{r Install Packages, echo=TRUE, message=FALSE, warning=FALSE}
## formatR - Required to format R code in the markdown ----
if (!is.element("formatR", installed.packages()[, 1])) {
  install.packages("formatR", dependencies = TRUE,
                   repos="https://cloud.r-project.org")
}
require("formatR")


## readr - Load datasets from CSV files ----
if (!is.element("readr", installed.packages()[, 1])) {
  install.packages("readr", dependencies = TRUE,
                   repos="https://cloud.r-project.org")
}
require("readr")

## MICE ----
# We use the MICE package to perform data imputation
if (!is.element("mice", installed.packages()[, 1])) {
  install.packages("mice", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
require("mice")
## dplyr ----
if (!is.element("dplyr", installed.packages()[, 1])) {
  install.packages("dplyr", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
require("dplyr")
#naniar ----
if (!is.element("naniar", installed.packages()[, 1])) {
  install.packages("naniar", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
require("naniar")
# Amelia ----
if (!is.element("Amelia", installed.packages()[, 1])) {
  install.packages("Amelia", dependencies = TRUE)
}
require("Amelia")
# corrplot ----
if (!is.element("corrplot", installed.packages()[, 1])) {
  install.packages("corrplot", dependencies = TRUE)
}
require("corrplot")
## ggplot2 ----
if (require("ggplot2")) {
  require("ggplot2")
} else {
  install.packages("ggplot2", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## caret ----
if (require("caret")) {
  require("caret")
} else {
  install.packages("caret", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## mlbench ----
if (require("mlbench")) {
  require("mlbench")
} else {
  install.packages("mlbench", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## pROC ----
if (require("pROC")) {
  require("pROC")
} else {
  install.packages("pROC", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
# language server ----
if (require("languageserver")) {
  require("languageserver")
} else {
  install.packages("languageserver", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
if (!is.element("ggcorrplot", installed.packages()[, 1])) {
  install.packages("ggcorrplot", dependencies = TRUE)
}
require("ggcorrplot")
## e1071 ----
if (require("e1071")) {
  require("e1071")
} else {
  install.packages("e1071", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## factoextra ----
if (require("factoextra")) {
  require("factoextra")
} else {
  install.packages("factoextra", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## FactoMineR ----
if (require("FactoMineR")) {
  require("FactoMineR")
} else {
  install.packages("FactoMineR", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## kernlab ----
if (require("kernlab")) {
  require("kernlab")
} else {
  install.packages("kernlab", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## rpart ----
if (require("rpart")) {
  require("rpart")
} else {
  install.packages("rpart", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
```

------------------------------------------------------------------------

**Note:** the following "*KnitR*" options have been set as the defaults in this markdown:\
`knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy.opts = list(width.cutoff = 80), tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	eval = TRUE,
	echo = TRUE,
	warning = FALSE,
	collapse = FALSE,
	tidy = TRUE
)
```

------------------------------------------------------------------------

**Note:** the following "*R Markdown*" options have been set as the defaults in this markdown:

> output:\
> \
> github_document:\
> toc: yes\
> toc_depth: 4\
> fig_width: 6\
> fig_height: 4\
> df_print: default\
> \
> editor_options:\
> chunk_output_type: console


# Loading the gum disease Dataset

The gum disease datatset is then loaded. The dataset and its metadata are available here: <https://drive.google.com/drive/folders/1-BGEhfOwquXF6KKXwcvrx7WuZXuqmW9q?usp=sharing>


```{r Your 1 Code Chunk}
library(readr)
Data_of_teeth <- read_csv("../data/Data of teeth.csv")

```

# Explatory data analysis 
## Dimensions 
 Dimensions refer to the number of observations (rows) and the number of
 attributes/variables/features (columns). Execute the following commands to
 display the dimensions of your datasets:
```{r Your 2 Code Chunk}
dim(Data_of_teeth)
```

## Data Types 
 Knowing the data types will help you to identify the most appropriate
 visualization types and algorithms that can be applied. It can also help you
 to identify the need to convert from categorical data (factors) to integers
 or vice versa where necessary. Execute the following command to identify the
 data types:
```{r Your 3 Code Chunk}
sapply(Data_of_teeth, class)
```

# Descriptive Statistics 
## Measures of Frequency 
```{r Your 4 Code Chunk}
disease_density_freq <- Data_of_teeth$Disease
cbind(frequency = table(disease_density_freq), percentage = prop.table(table(disease_density_freq)) *
        100)

symptom1_density_freq <- Data_of_teeth$"Symptom 1"
cbind(frequency = table(symptom1_density_freq), percentage = prop.table(table(symptom1_density_freq)) *
        100)

symptom2_density_freq <- Data_of_teeth$"Symptom 2"
cbind(frequency = table(symptom2_density_freq), percentage = prop.table(table(symptom2_density_freq)) *
        100)
symptom3_density_freq <- Data_of_teeth$"Symptom 3"
cbind(frequency = table(symptom3_density_freq), percentage = prop.table(table(symptom3_density_freq)) *
        100)
treatment_density_freq <- Data_of_teeth$Treatment
cbind(frequency = table(treatment_density_freq), percentage = prop.table(table(treatment_density_freq)) *
        100)
```


## Measures of Distribution/Dispersion/Spread/Scatter/Variability 
```{r Your 5 Code Chunk}
summary(Data_of_teeth)
```

## changing the symptom into numeric values
```{r Your 6 Code Chunk}
# Coding scheme for symptoms
symptom_coding <- c(
  "a cracked tooth" = 1, 
  "Bad breath" = 2, 
  "Black, white, or brown tooth stains" = 3, 
  "bleeding" = 4, 
  "Cracked or chipped teeth" = 5, 
  "Difficulty chewing or swallowing" = 6, 
  "Dramatic weight loss" = 7, 
  "Ear Pain" = 8, 
  "Grooves on your teeth’s surface" = 9, 
  "gum disease" = 10, 
  "Holes or pits in your teeth" = 11, 
  "pain" = 12, 
  "Pain when you bite down" = 13, 
  "Painful chewing" = 14, 
  "Red and swollen gums" = 15, 
  "sore throat" = 16, 
  "Tender or bleeding gums" = 17, 
  "worn-down fillings or crowns" = 18, 
  "Yellowish discoloration" = 19
)

# Applying coding scheme to create a new column
teeth_Num <- Data_of_teeth
teeth_Num$symptom1 <- symptom_coding[Data_of_teeth$"Symptom 1"]
teeth_Num$symptom2 <- symptom_coding[Data_of_teeth$"Symptom 2"]
teeth_Num$symptom3 <- symptom_coding[Data_of_teeth$"Symptom 3"]
```

## univariate plots
### Create Box and Whisker Plots for Each Numeric Attribute 
```{r Your 7 Code Chunk}
boxplot(teeth_Num[, 7], main = names(teeth_Num)[7])
boxplot(teeth_Num[, 8], main = names(teeth_Num)[8])
boxplot(teeth_Num[, 9], main = names(teeth_Num)[9])
```

###  Create Bar Plots for Each Categorical Attribute 
```{r Your 8 Code Chunk}
barplot(table(teeth_Num[, 4]), main = names(teeth_Num)[4])

```

###  Create a Missingness Map to Identify Missing Data 
```{r Your 9 Code Chunk}
missmap(Data_of_teeth, col = c("red", "grey"), legend = TRUE)

```

## multivariate plots
### Create a Correlation Plot 
 Correlation plots can be used to get an idea of which attributes change
 together. The function “corrplot()” found in the package “corrplot” is
 required to perform this. The larger the dot in the correlation plot, the
 larger the correlation. Blue represents a positive correlation whereas red
 represents a negative correlation.
```{r Your 10 Code Chunk}
corrplot(cor(teeth_Num[, 7:9]), method = "circle")
```

 Alternatively, the 'ggcorrplot::ggcorrplot()' function can be used to plot a
 more visually appealing plot.
```{r Your 11 Code Chunk}
ggcorrplot(cor(teeth_Num[, 7:9]))
```

# preprocessing and data transformation
## Confirm the "missingness" in the Dataset before Imputation 
 Are there missing values in the dataset?
```{r Your 12 Code Chunk}
any_na(Data_of_teeth)
any_na(teeth_Num)
``` 


How many?
```{r Your 13 Code Chunk}
n_miss(Data_of_teeth)
n_miss(teeth_Num)
```


What is the percentage of missing data in the entire dataset?
```{r Your 14 Code Chunk}
prop_miss(Data_of_teeth)
prop_miss(teeth_Num)
```


How many missing values does each variable have?
```{r Your 15 Code Chunk}
Data_of_teeth %>% is.na() %>% colSums()
teeth_Num %>% is.na() %>% colSums()
```


 What is the number and percentage of missing values grouped by
 each variable?
```{r Your 16 Code Chunk}
miss_var_summary(Data_of_teeth)
miss_var_summary(teeth_Num)
```


remove the column that has misssing data
```{r Your 17 Code Chunk}
Data_of_teeth<-Data_of_teeth[-6]
teeth_Num<-teeth_Num[-6]
```


 remove the symptom columns that have character values
```{r Your 18 Code Chunk}
teeth_Num<-teeth_Num[,-1:-3]

```

 remove the treatment columns that have character values
```{r Your 19 Code Chunk}
teeth_Num<-teeth_Num[-2]
```


confirm if there is missing data
```{r Your 20 Code Chunk}
any_na(Data_of_teeth)
any_na(teeth_Num)
```

# Selection of model
## Split the dataset 
 Define a 75:25 train:test data split of the dataset.
 That is, 75% of the original data will be used to train the model and 25% of the original data will be used to test the model.
```{r Your 212 Code Chunk}
train_index <- createDataPartition(teeth_Num$Disease,
                                   p = 0.75,
                                   list = FALSE)
teeth_disease_train <- teeth_Num[train_index, ]
teeth_disease_test <- teeth_Num[-train_index, ]
```


## 1. Decision tree for a classification problem with caret 
 
```{r Your 21 Code Chunk}
### Train the model
set.seed(7)
# We apply the 5-fold cross validation resampling method
train_control <- trainControl(method = "cv", number = 5)
teeth_caret_model_rpart <- train(Disease ~ ., data = teeth_Num,
                                    method = "rpart", metric = "Accuracy",
                                    trControl = train_control)

# Display the model's details 
print(teeth_caret_model_rpart)

# Make predictions 
predictions <- predict(teeth_caret_model_rpart,
                       teeth_disease_test[, 2:4],
                       type = "raw")

# Display the model's evaluation metrics 
table(predictions, teeth_disease_test$Disease)

confusion_matrix <-
  caret::confusionMatrix(predictions,
                         as.factor(teeth_disease_test[, 1:4]$Disease))
print(confusion_matrix)
```


## 2.  Naïve Bayes 
```{r Your 22 Code Chunk}
#### Train the model ----
# We apply the 5-fold cross validation resampling method
set.seed(7)
train_control <- trainControl(method = "cv", number = 5)
teeth_caret_model_nb <- train(Disease ~ .,
                                 data = teeth_disease_train,
                                 method = "nb", metric = "Accuracy",
                                 trControl = train_control)

#### Display the model's details ----
print(teeth_caret_model_nb)

#### Make predictions ----
predictions <- predict(teeth_caret_model_nb,
                       teeth_disease_test[, 2:4])

#### Display the model's evaluation metrics ----
confusion_matrix <-
  caret::confusionMatrix(predictions,
                         as.factor(teeth_disease_test[, 1:4]$Disease))
print(confusion_matrix)
```


## 3. kNN for a classification problem with CARET's train function  
```{r Your 23 Code Chunk}
set.seed(7)
train_control <- trainControl(method = "cv", number = 10)
teeth_caret_model_knn <- train(Disease ~ ., data = teeth_Num,
                                  method = "knn", metric = "Accuracy",
                                  preProcess = c("center", "scale"),
                                  trControl = train_control)

#### Display the model's details ----
print(teeth_caret_model_knn)

#### Make predictions ----
predictions <- predict(teeth_caret_model_knn,
                       teeth_disease_test[, 2:4])

#### Display the model's evaluation metrics ----
confusion_matrix <-
  caret::confusionMatrix(predictions,
                         as.factor(teeth_disease_test[, 1:4]$Disease))
print(confusion_matrix)
```


## 4. SVM Classifier for a classification problem with CARET 
```{r Your 24 Code Chunk}
#### Train the model ----
set.seed(7)
train_control <- trainControl(method = "cv", number = 5)
teeth_caret_model_svm_radial <- # nolint: object_length_linter.
  train(Disease ~ ., data = teeth_disease_train, method = "svmRadial",
        metric = "Accuracy", trControl = train_control)

#### Display the model's details ----
print(teeth_caret_model_svm_radial)

#### Make predictions ----
predictions <- predict(teeth_caret_model_svm_radial,
                       teeth_disease_test[, 2:4])

#### Display the model's evaluation metrics ----
table(predictions, teeth_disease_test$Disease)
confusion_matrix <-
  caret::confusionMatrix(predictions,
                         as.factor(teeth_disease_test[, 1:4]$Disease))
print(confusion_matrix)
```

#model performance comparison 
```{r Your 244 Code Chunk}
train_control <- trainControl(method = "repeatedcv", number = 10, repeats = 3)

### CART ----
set.seed(7)
teeth_model_cart <- train(Disease ~ ., data = teeth_Num,
                             method = "rpart", trControl = train_control)

### KNN ----
set.seed(7)
teeth_model_knn <- train(Disease ~ ., data = teeth_Num,
                            method = "knn", trControl = train_control)

### SVM ----
set.seed(7)
teeth_model_svm <- train(Disease ~ ., data = teeth_Num,
                            method = "svmRadial", trControl = train_control)

### naive bayes ----
set.seed(7)
teeth_model_nb <- train(Disease ~ ., data = teeth_Num,
                         method = "nb", trControl = train_control)
```

##  Call the `resamples` Function 
We then create a list of the model results and pass the list as an argument
to the `resamples` function.
```{r Your 25 Code Chunk}
results <- resamples(list( CART = teeth_model_cart,
                          KNN = teeth_model_knn, SVM = teeth_model_svm,
                          NB = teeth_model_nb))
```



##  Display the Results 
### 1. Table Summary 
 This is the simplest comparison. It creates a table with one model per row
 and its corresponding evaluation metrics displayed per column.
```{r Your 26 Code Chunk}
summary(results)
```


### 2. Box and Whisker Plot 
 This is useful for visually observing the spread of the estimated accuracies
 for different algorithms and how they relate.
```{r Your 266 Code Chunk}
scales <- list(x = list(relation = "free"), y = list(relation = "free"))
bwplot(results, scales = scales)
```


### 3. Dot Plots 
 They show both the mean estimated accuracy as well as the 95% confidence
 interval (e.g. the range in which 95% of observed scores fell).
```{r Your 27 Code Chunk}
scales <- list(x = list(relation = "free"), y = list(relation = "free"))
dotplot(results, scales = scales)
```


### 4. Scatter Plot Matrix 
 This is useful when considering whether the predictions from two
 different algorithms are correlated. If weakly correlated, then they are good
 candidates for being combined in an ensemble prediction.
```{r Your 277 Code Chunk}
splom(results)
```


### 5. Pairwise xyPlots 
 You can zoom in on one pairwise comparison of the accuracy of trial-folds for
 two models using an xyplot.
```{r Your 28 Code Chunk}
# xyplot plots to compare models
xyplot(results, models = c("KNN", "SVM"))


# xyplot plots to compare models
xyplot(results, models = c("SVM", "CART"))


# xyplot plots to compare models
xyplot(results, models = c("KNN", "CART"))


# xyplot plots to compare models
xyplot(results, models = c("KNN", "NB"))

# or
# xyplot plots to compare models
xyplot(results, models = c("NB", "CART"))
```


### 6. Statistical Significance Tests 
 This is used to calculate the significance of the differences between the
 metric distributions of the various models.

 Upper Diagonal 
 The upper diagonal of the table shows the estimated difference between the
 distributions. If we think that LDA is the most accurate model from looking
 at the previous graphs, we can get an estimate of how much better it is than
 specific other models in terms of absolute accuracy.

 Lower Diagonal 
 The lower diagonal contains p-values of the null hypothesis.
 The null hypothesis is a claim that "the distributions are the same".
 A lower p-value is better (more significant).
```{r Your 29 Code Chunk}
diffs <- diff(results)

summary(diffs)
```



# \<You can Provide Another Appropriate Title Here if you wish\>
```{r Your 0 Code Chunk}
# yes working
```
Describe the code chunk here:

```{r Your Sixth Code Chunk}
# Fill this with R related code that will be executed when the R markdown file
```

## \<You Can Have a Sub-Title Here if you wish\>

```{r Your Seventh Code Chunk}
# Fill this with other R related code that will be executed when the R markdown
```

## \<You Can Have Another Sub-Title Here if you wish\>

```{r Your Eighth Code Chunk}
# Fill this with other R related code that will be executed when the R markdown
```

**etc.** as per the lab submission requirements. Be neat and communicate in a clear and logical manner.