---
title: "Business Intelligence Lab Submission Markdown"
author: "<trevor ngugi>"
date: "<27/11/2023>"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Student ID Numbers and Names of Group Members** | |
|                                                   |                                                                                                                                                                          |
|                                                   | 1.  133996 - B - Trevor ngugi                                                                                                                                              |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                               |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                             |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                               |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                               |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **GitHub Classroom Group Name**                   | champions                                                                                                       |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Course Code**                                   | BBT4206                                                                                                                                                                  |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Course Name**                                   | Business Intelligence II                                                                                                                                                 |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Program**                                       | Bachelor of Business Information Technology                                                                                                                              |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Semester Duration**                             | 21^st^ August 2023 to 28^th^ November 2023                                                                                                                               |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

# Setup Chunk

We start by installing all the required packages

```{r Install Packages, echo=TRUE, message=FALSE, warning=FALSE}
## formatR - Required to format R code in the markdown ----
if (!is.element("formatR", installed.packages()[, 1])) {
  install.packages("formatR", dependencies = TRUE,
                   repos="https://cloud.r-project.org")
}
require("formatR")


## readr - Load datasets from CSV files ----
if (!is.element("readr", installed.packages()[, 1])) {
  install.packages("readr", dependencies = TRUE,
                   repos="https://cloud.r-project.org")
}
require("readr")

## MICE ----
# We use the MICE package to perform data imputation
if (!is.element("mice", installed.packages()[, 1])) {
  install.packages("mice", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
require("mice")
## dplyr ----
if (!is.element("dplyr", installed.packages()[, 1])) {
  install.packages("dplyr", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
require("dplyr")
#naniar ----
if (!is.element("naniar", installed.packages()[, 1])) {
  install.packages("naniar", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
require("naniar")
# Amelia ----
if (!is.element("Amelia", installed.packages()[, 1])) {
  install.packages("Amelia", dependencies = TRUE)
}
require("Amelia")
# corrplot ----
if (!is.element("corrplot", installed.packages()[, 1])) {
  install.packages("corrplot", dependencies = TRUE)
}
require("corrplot")
## ggplot2 ----
if (require("ggplot2")) {
  require("ggplot2")
} else {
  install.packages("ggplot2", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## caret ----
if (require("caret")) {
  require("caret")
} else {
  install.packages("caret", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## mlbench ----
if (require("mlbench")) {
  require("mlbench")
} else {
  install.packages("mlbench", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## pROC ----
if (require("pROC")) {
  require("pROC")
} else {
  install.packages("pROC", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
# language server ----
if (require("languageserver")) {
  require("languageserver")
} else {
  install.packages("languageserver", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
if (!is.element("ggcorrplot", installed.packages()[, 1])) {
  install.packages("ggcorrplot", dependencies = TRUE)
}
require("ggcorrplot")
```

------------------------------------------------------------------------

**Note:** the following "*KnitR*" options have been set as the defaults in this markdown:\
`knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy.opts = list(width.cutoff = 80), tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	eval = TRUE,
	echo = TRUE,
	warning = FALSE,
	collapse = FALSE,
	tidy = TRUE
)
```

------------------------------------------------------------------------

**Note:** the following "*R Markdown*" options have been set as the defaults in this markdown:

> output:\
> \
> github_document:\
> toc: yes\
> toc_depth: 4\
> fig_width: 6\
> fig_height: 4\
> df_print: default\
> \
> editor_options:\
> chunk_output_type: console


# Loading the gum disease Dataset

The gum disease datatset is then loaded. The dataset and its metadata are available here: <https://drive.google.com/drive/folders/1-BGEhfOwquXF6KKXwcvrx7WuZXuqmW9q?usp=sharing>


```{r Your 1 Code Chunk}
library(readr)
Data_of_teeth <- read_csv("../data/Data of teeth.csv")

```

# Explatory data analysis 
## Dimensions 
 Dimensions refer to the number of observations (rows) and the number of
 attributes/variables/features (columns). Execute the following commands to
 display the dimensions of your datasets:
```{r Your 2 Code Chunk}
dim(Data_of_teeth)
```

## Data Types 
 Knowing the data types will help you to identify the most appropriate
 visualization types and algorithms that can be applied. It can also help you
 to identify the need to convert from categorical data (factors) to integers
 or vice versa where necessary. Execute the following command to identify the
 data types:
```{r Your 3 Code Chunk}
sapply(Data_of_teeth, class)
```

# Descriptive Statistics 
## Measures of Frequency 
```{r Your 4 Code Chunk}
disease_density_freq <- Data_of_teeth$Disease
cbind(frequency = table(disease_density_freq), percentage = prop.table(table(disease_density_freq)) *
        100)

symptom1_density_freq <- Data_of_teeth$"Symptom 1"
cbind(frequency = table(symptom1_density_freq), percentage = prop.table(table(symptom1_density_freq)) *
        100)

symptom2_density_freq <- Data_of_teeth$"Symptom 2"
cbind(frequency = table(symptom2_density_freq), percentage = prop.table(table(symptom2_density_freq)) *
        100)
symptom3_density_freq <- Data_of_teeth$"Symptom 3"
cbind(frequency = table(symptom3_density_freq), percentage = prop.table(table(symptom3_density_freq)) *
        100)
treatment_density_freq <- Data_of_teeth$Treatment
cbind(frequency = table(treatment_density_freq), percentage = prop.table(table(treatment_density_freq)) *
        100)
```


## Measures of Distribution/Dispersion/Spread/Scatter/Variability 
```{r Your 5 Code Chunk}
summary(Data_of_teeth)
```

## changing the symptom into numeric values
```{r Your 6 Code Chunk}
# Coding scheme for symptoms
symptom_coding <- c(
  "a cracked tooth" = 1, 
  "Bad breath" = 2, 
  "Black, white, or brown tooth stains" = 3, 
  "bleeding" = 4, 
  "Cracked or chipped teeth" = 5, 
  "Difficulty chewing or swallowing" = 6, 
  "Dramatic weight loss" = 7, 
  "Ear Pain" = 8, 
  "Grooves on your teeth’s surface" = 9, 
  "gum disease" = 10, 
  "Holes or pits in your teeth" = 11, 
  "pain" = 12, 
  "Pain when you bite down" = 13, 
  "Painful chewing" = 14, 
  "Red and swollen gums" = 15, 
  "sore throat" = 16, 
  "Tender or bleeding gums" = 17, 
  "worn-down fillings or crowns" = 18, 
  "Yellowish discoloration" = 19
)

# Applying coding scheme to create a new column
teeth_Num <- Data_of_teeth
teeth_Num$symptom1 <- symptom_coding[Data_of_teeth$"Symptom 1"]
teeth_Num$symptom2 <- symptom_coding[Data_of_teeth$"Symptom 2"]
teeth_Num$symptom3 <- symptom_coding[Data_of_teeth$"Symptom 3"]
```

## univariate plots
### Create Box and Whisker Plots for Each Numeric Attribute 
```{r Your 7 Code Chunk}
boxplot(teeth_Num[, 7], main = names(teeth_Num)[7])
boxplot(teeth_Num[, 8], main = names(teeth_Num)[8])
boxplot(teeth_Num[, 9], main = names(teeth_Num)[9])
```

###  Create Bar Plots for Each Categorical Attribute ----
```{r Your 8 Code Chunk}
barplot(table(teeth_Num[, 4]), main = names(teeth_Num)[4])

```

###  Create a Missingness Map to Identify Missing Data ----
```{r Your 9 Code Chunk}
missmap(Data_of_teeth, col = c("red", "grey"), legend = TRUE)

```

## multivariate plots
### Create a Correlation Plot ----
 Correlation plots can be used to get an idea of which attributes change
 together. The function “corrplot()” found in the package “corrplot” is
 required to perform this. The larger the dot in the correlation plot, the
 larger the correlation. Blue represents a positive correlation whereas red
 represents a negative correlation.
```{r Your 10 Code Chunk}
corrplot(cor(teeth_Num[, 7:9]), method = "circle")
```

 Alternatively, the 'ggcorrplot::ggcorrplot()' function can be used to plot a
 more visually appealing plot.
```{r Your 11 Code Chunk}
ggcorrplot(cor(teeth_Num[, 7:9]))
```

# \<You can Provide Another Appropriate Title Here if you wish\>
```{r Your 0 Code Chunk}
# yes working
```
Describe the code chunk here:

```{r Your Sixth Code Chunk}
# Fill this with R related code that will be executed when the R markdown file
```

## \<You Can Have a Sub-Title Here if you wish\>

```{r Your Seventh Code Chunk}
# Fill this with other R related code that will be executed when the R markdown
```

## \<You Can Have Another Sub-Title Here if you wish\>

```{r Your Eighth Code Chunk}
# Fill this with other R related code that will be executed when the R markdown
```

**etc.** as per the lab submission requirements. Be neat and communicate in a clear and logical manner.